maze-dataset

### 2025-03-30 docs/paper/paper.tex
suggestions from @bknutson0

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this very fact makes it an excellent testbed for studying how
machine learning algorithms solve problems and represent spatial
information. We introduce \texttt{maze-dataset}, a Python library for
generating, processing, and visualizing datasets of mazes. This library
supports a variety of maze generation algorithms providing both mazes
with loops and ``perfect'' mazes without them. These generation
algorithms can be configured with various parameters, and the resulting
mazes can be filtered to satisfy desired properties. Also provided are
tools for converting mazes to and from various formats suitable for a
variety of neural network architectures, such as rasterized images and
tokenized text sequences, as well as various visualization tools. As
well as providing a simple interface for generating, storing, and
loading these datasets, \texttt{maze-dataset} is extensively tested,
type hinted, benchmarked, and documented.

@@

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties. Custom filters can be specified, and some filters are
included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.

@@

This package utilizes a simple, efficient representation of mazes. Using
an adjacency list to represent mazes would lead to a poor lookup time of
whether any given connection exists, while using an adjacency matrix
would waste memory by failing to exploit the structure (e.g., only 4 of
the diagonals would be filled in). Instead, we describe mazes with the
following simple representation: for a \(d\)-dimensional lattice with
\(r\) rows and \(c\) columns, we initialize a boolean array
\(A = \{0, 1\}^{d \times r \times c}\), which we refer to in the code as
a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a downward connection
exists from node \([i,j]\) to \([i+1, j]\). Likewise, the value at
\(A[1,i,j]\) determines whether a rightwards connection to \([i, j+1]\)
exists. Thus, we avoid duplication of data about the existence of
connections, at the cost of requiring additional care with indexing when
looking for a connection upwards or to the left. Note that this setup
allows for a periodic lattice.

@@

\begin{itemize}
\item
  (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) use
  \texttt{maze-dataset} to compare the effectiveness of transformers
  trained with the MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) use
  \texttt{maze-dataset} to study the effectiveness of imperative
  learning
\end{itemize}

>>>>

Solving mazes is a classic problem in computer science and artificial
intelligence, and humans have been constructing mazes for thousands of
years. Although finding the shortest path through a maze is a solved
problem, this very fact makes it an excellent testbed for studying how
machine learning algorithms solve problems and represent spatial
information. We introduce \texttt{maze-dataset}, a Python library for
generating, processing, and visualizing datasets of mazes. This library
supports a variety of maze generation algorithms providing mazes with or
without loops, fully-connected or not, and many other variations. These
generation algorithms can be configured with various parameters, and the
resulting mazes can be filtered to satisfy desired properties. Also
provided are tools for converting mazes to and from various formats
suitable for a variety of neural network architectures, such as
rasterized images, tokenized text sequences, and various visualizations.
As well as providing a simple interface for generating, storing, and
loading these datasets, \texttt{maze-dataset} is extensively tested,
type hinted, benchmarked, and documented.

@@

Furthermore, a dataset of mazes can be filtered to satisfy certain
properties. Custom filters can be specified, and some filters are
included in
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/filters.html\#MazeDatasetFilters}{\texttt{MazeDatasetFilters}}.
For example, we can require a minimum path length from origin to target
of 3:

@@

This package utilizes a simple, efficient representation of mazes. Using
an adjacency list to represent mazes would lead to a poor lookup time
for whether any given connection exists, while using an adjacency matrix
would waste memory by failing to exploit the structure (e.g., only 4 of
the diagonals would be filled in). Instead, we describe mazes with the
following simple representation: for a \(d\)-dimensional lattice with
\(r\) rows and \(c\) columns, we initialize a boolean array
\(A = \{0, 1\}^{d \times r \times c}\), which we refer to in the code as
a
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#LatticeMaze.connection_list}{\texttt{connection\_list}}.
The value at \(A[0,i,j]\) determines whether a downward connection
exists from node \([i,j]\) to \([i+1, j]\). Likewise, the value at
\(A[1,i,j]\) determines whether a rightwards connection to \([i, j+1]\)
exists. Thus, we avoid duplication of data about the existence of
connections, at the cost of requiring additional care with indexing when
looking for a connection upwards or to the left. Note that this setup
allows for a periodic lattice.

@@

\begin{itemize}
\item
  (\protect\hyperlink{ref-nolte2024multistep}{Nolte et al., 2024}) use
  \texttt{maze-dataset} to compare the effectiveness of transformers
  trained with the MLM-\(\mathcal{U}\)
  (\protect\hyperlink{ref-MLMU-kitouni2024factorization}{Kitouni et al.,
  2024}) multistep prediction objective against standard autoregressive
  training for multi-step planning on our maze task.
\item
  (\protect\hyperlink{ref-wang2024imperative}{Wang et al., 2024}) and
  (\protect\hyperlink{ref-chen2024iaimperative}{Chen et al., 2024}) use
  \texttt{maze-dataset} to study the effectiveness of imperative
  learning.
\end{itemize}
